A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-09-23 main.tex
3.6.4 edits

\subsubsection{Structure of environmental effects}

\pb{Perhaps ``Patterns of environmental effects'' would be better?}

For each $\bX$ matrix, $\bZ$ was a corresponding incidence matrix of 1s and 0s identifying subpopulation, or founding population membership \pb{you don't need to keep saying ``or founding population'' over and over again}, which assigns the corresponding environmental effect found in $\bgamma$. Various $\bgamma$ structures were considered in order to investigate the effects of various forms of environmental confounding.

Our initial hypothesis was that structures of $\bgamma$ that can be well-approximated by a linear combination of the observed genetic data would be well-approximated and corrected for by PC-Lasso, whereas LMM-Lasso may be more able to correct for subpopulation-specific effects that are nonlinear in nature and as such do not lend themselves to such an approximation \pb{is the ``linearity'' the key issue here?  I'd say it's more a matter of complexity: PC-Lasso would work for simple patterns, but can't represent more complex patterns}. To that end we investigated two forms of environmental effect magnitudes: (1) exponential, and (2) dichotomous \pb{neither of which are linear, so this doesn't transition well}. In addition to the magnitudes of effects, we also considered whether individuals in genetically similar subpopulations would experience similar environmental effects \pb{Shouldn't we lead with this?  I mean, isn't this the most important concept in this section?}. For example, a geographic environmental exposure such as the air quality described in our motivating example is likely to be similar among genetically related individuals due to concordance between genetic clusters and geography. However, behavior-based environmental exposures may be less likely to be shared by subpopulations which are genetically and geographically similar. As an example, \citet{ng2014smoking} found considerable within-region variation in smoking rates in Africa, Asia, and South America.

\anna{complex/noncomplex ... can be complex in different ways... discordant vs concordant ... dichotomous vs exponential}

 To evaluate the performance of penalized LMM methods in both scenarios, we consider exponential and dichotomous effect magnitudes with are concordant or discordant in nature, for a total of four distinct $\bgamma$ structures. We define concordant environmental effects as those that are similar among genetically similar subpopulations, and discordant environmental effects as those that are not. Figure \ref{fig:gamma_structures} provides a visual summary of the four coefficient structures. 1000 simulations were performed for each unique combination of parameters, datatype, subpopulation structure, and environmental effect structure. \pb{need to describe somehow that subpopulations are ordered with respect to how similar they are?}
\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/gamma_structure.png}
    \caption{$\bgamma$ structures \pb{probably want a better caption}}
    \label{fig:gamma_structures}
\end{figure}

@@

>>>>

\subsubsection{Patterns of environmental effects}

For each $\bX$ matrix, $\bZ$ was a corresponding incidence matrix of 1s and 0s identifying subpopulation, or founding population membership, which assigns the corresponding environmental effect found in $\bgamma$. Various $\bgamma$ patterns were considered in order to investigate the effects of distinct forms of environmental confounding. Our initial hypothesis was that simple patterns of environmental effects could be well-approximated by a linear combination of the observed genetic data and well-corrected for by both PC-Lasso and LMM-Lasso, while PC-Lasso would be unable to represent and correct for more complex patterns of environmental effects.

Environmental effects may be complex in different ways. Firstly, we considered whether individuals in genetically similar subpopulations would experience similar environmental effects. For example, a geographic environmental exposure such as the air quality described in our motivating example is likely to be similar among genetically related individuals due to concordance between genetic clusters and geography. However, behavior-based environmental exposures may be less likely to be shared by subpopulations which are genetically and geographically similar. As an example, \citet{ng2014smoking} found considerable within-region variation in smoking rates in Africa, Asia, and South America. We refer to environmental effects that are similar among genetically similar subpopulations as concordant, and to those that are not as discordant. Secondly, we considered two patterns of environmental effect sizes, which we refer to as exponential and dichotomous. We use the term dichotomous to describe two effect sizes each experienced by half of the population, regardless of the number of subpopulations. While exponential effects refer to a unique effect size for each subpopulation with exponentially increasing magnitudes.

 To evaluate the performance of penalized LMM methods in the presence of these different forms of confounding complexity, we consider environmental effects which are concordant or discordant in nature in the presence of both exponential and dichotomous effect sizes, for a total of four distinct $\bgamma$ patterns. Figure \ref{fig:gamma_structures} provides a visual summary of the four coefficient patterns, where the subpopulations on the x-axis are ordered by genetic similarity so that subpopulation 1 is most genetically similar to subpopulation 2, and least similar to subpopulation 10. This may also be thought of as subpopulations ordered spatially along a 1-dimensional geographic region.
\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/gamma_structure.png}
    \caption{An illustration of four distinct patterns of environmental effects and corresponding $\bgamma$ values in the presence of 10 subpopulations ordered by genetic similarity.}
    \label{fig:gamma_structures}
\end{figure}
\noindent In the case of concordant and dichotomous effects, Figure \ref{fig:gamma_structures} shows that the first five ordered subpopulations are subject to an effect size of 1, while the second five are subject to an effect size of $-$1. In the case of discordant dichotomous effects, the first five subpopulations experience an effect size with an absolute value of \anna{1}, and the second five an absolute effect size of \anna{0.5}. However, the directions of these absolute effects sizes alternate among the ordered subpopulations so that genetically similar subpopulations do not experience similar environmental effects. \anna{not quite 1 and 0.5 because of centering and standardization...how to word this so it's not confusing?}

@@

1000 simulations were performed for each unique combination of parameters, datatype, subpopulation structure, and environmental effect structure.
